# Figures 6 and 7

```{r imports}
#| cache: false
#| message: false
library(dplyr)
library(Matrix) # spMatrix
library(Rglpk)  # Rglpk_solve_LP
library(readr)
library(stringr)
library(tidyr)
library(forcats)
library(ggplot2)
library(EBImage)

source("R/config.R")
```

## Toy example of digital partners

Use a linear programming solver in R to run unbalanced optimal transport (OT).
Although more low-level than a framework like Python OT, linear programming has
the flexibility to setup our problem where only certain GranSim simulations
being matched against MIBI granulomas instead of an all-to-all comparison.

```{r pair-toy}
#| cache: true
## Toy example of 2 MIBI granulomas and 3 GranSim granulomas for each MIBI.
##
## Nevertheless there are some challenges:
##
## 1. We have a feature vector instead of a single value for each i and j
##    observation.  We reduce this dimension when setting up the cost matrix as
##    described in equation 3 of the methods.
##
## 2. We must limit a given MIBI granuloma to only be compared against the
##    subset of GranSim granulomas.  This is addressed by fixing values of the
##    output x_{i,j} to 0 as described in the methods.
##
## Please note that this implementation avoids the 3-dimensional matrices G'
## and M' of equation 3 by instead using an easier to understand 2 dimensional
## ordering of G and M.  Part of the reason for using this 2-dimensional
## pairwise approach and not matching equation 3 is the trouble using
## crossprod() in R for 3-dimensional arrays, but perhaps it also keeps the
## code simpler and easier to debug.

n_feat <- 4 # Symbolized by n(F) in the methods equation 3.
## MIBI feature matrix, symbolized by M with dimensions [F, D] in the methods.
mat_mibi <-
  matrix(c(1, 2, 3, 4,
           5, 6, 7, 8),
         ncol = n_feat,
         byrow = TRUE) %>%
  t()
mat_mibi
## Symbolized by n(D) in the methods all equations.
n_mibi <- ncol(mat_mibi)
## Number of GranSim granulomas per MIBI granuloma.  No explicit symbol for
## this in the methods, but a property of how our data is ordered that helps us
## setup the constraints.
mat_gs_per_mibi <- c(3, 3)
stopifnot(length(mat_gs_per_mibi) == n_mibi)
## GranSim feature matrix, symbolized by G with dimensions [F, O] in the
## methods.
mat_gs <-
  matrix(c(1, 2, 3, 4,
           0, 1, 2, 3,
           2, 3, 4, 5,
           4, 5, 6, 7,
           5, 6, 7, 8,
           6, 7, 8, 9),
         ncol = n_feat,
         byrow = TRUE) %>%
  t()
mat_gs
## >= n(O) in the methods all equations, because it may contain repeated
## GranSim granulomas to match the MIBI granulomas.
n_gransim <- ncol(mat_gs)
## Symbolized by E in the methods equation 3.
elpd <- c(30, 20, 10, 15)
## Reduce the ELPD and feature matrices into the cost matrix symbolized by c,
## as explained in the methods equation 3.
cost_vec <-
  elpd %*% abs(mat_gs - mat_mibi[, rep(seq_len(n_mibi),
                                       times = mat_gs_per_mibi)])
str(cost_vec)
## Encode the methods equation 2 constraint that only one mibi granuloma must
## be chosen from the set of all GranSim granulomas.
constraints <-
  spMatrix(nrow = n_mibi,
           ncol = n_gransim,
           i = rep(seq_len(n_mibi),
                   times = mat_gs_per_mibi),
           j = seq_len(n_gransim),
           x = rep(1, n_gransim))
constraints
ot <-
  Rglpk_solve_LP(obj = cost_vec,
                 mat = constraints,
                 dir = rep("==", nrow(constraints)),
                 rhs = rep(1, nrow(constraints)),
                 types = rep("B", length(cost_vec)))
str(ot)
gransim_idx <- c(sapply(mat_gs_per_mibi, seq_len))
str(gransim_idx)
df_pairs <-
  tibble(mibi = seq_len(n_mibi),
         gransim = gransim_idx[as.logical(ot$solution)])
df_pairs
```

## Actual digital partners plot

```{r prepare-inputs}
#| cache: true
file_elpd <-
  file.path(dir_data,
            "2024-09-03-A-varsel-50k",
            "output",
            "rank_week11.txt")
file_mibi <-
  file.path(dir_data,
            "2024-09-17-A-mibi-data-oneshot",
            "output-combined",
            "week11.csv")
file_gs <-
  file.path(dir_data,
            "2024-08-19-A-training-data-reduce-50k",
            "output-combined",
            "week11.csv")
file_gs_traj <-
  file.path(dir_data,
            "2024-07-19-A-gr-50k",
            "output-lhs-stat-cols-35.csv.gz")
cols_gs_all <-
  read_csv(file_gs,
           col_select = -1,
           n_max = 0L,
           show_col_types = FALSE) %>%
  colnames()
col_id <- cols_gs_all[1]
cols_gs <- c(
  col_id,
  "t_gam_AreaShape_BoundingBoxMaximum_Y_IQR",
  "mac_Neighbors_NumberOfNeighbors_3_mean",
  "macs_Intensity_StdIntensity_tnf_mean",
  "macs_RadialDistribution_FracAtD_ifng_3of4_mean",
  "ts_Intensity_StdIntensity_tnf_mean",
  "mac_Location_CenterMassIntensity_X_ifng_IQR",
  "t_gam_Children_t_gams_Count_mean",
  "mac_Location_CenterMassIntensity_Y_ifng_IQR",
  "t_cyts_AreaShape_EquivalentDiameter_mean",
  "t_regs_AreaShape_EquivalentDiameter_mean",
  "macs_Intensity_MaxIntensity_tnf_mean",
  "mac_Children_macs_Count_mean",
  "mac_Neighbors_FirstClosestDistance_t_reg_3_median",
  "gran_Intensity_StdIntensity_tnf_mean"
)
# Not all columns will be found in MIBI due to missing TNF, different promxity
# size (3[x 20um] vs 60[x 1um]), etc.
cols_mibi_all <-
  read_csv(file_mibi,
           col_select = -1,
           n_max = 0L,
           show_col_types = FALSE) %>%
  colnames()
cols <-
  intersect(cols_gs,
            str_replace(cols_mibi_all, "_60_", "_3_"))
cols_to_scale <-
  cols[-c(1, 2, 7)]

df_mibi <-
  read_csv(file_mibi,
           col_select = all_of(str_replace(cols, "_3_", "_60_")),
           show_col_types = FALSE) %>%
  ## Linear programming / mixed-integer programming does not handle NAs.
  drop_na() %>%
  arrange(get(col_id)) %>%
  rename(sample = 1) %>%
  mutate(sample = fct_inorder(str_c("sample", sample)))
df_gs <-
  read_csv(file_gs,
           col_select = all_of(cols),
           show_col_types = FALSE) %>%
  ## Linear programming / mixed-integer programming does not handle NAs.
  drop_na()  %>%
  arrange(get(col_id)) %>%
  rename(exp = 1) %>%
  mutate(exp = as.integer(exp)) %>%
  ## Rescale GranSim 20um measurements to MIBI-TOF 1um.
  mutate(across(all_of(cols_to_scale),
                function(x) x * 20))
df_elpd <-
  read_csv(file_elpd,
           col_select = c(col = "ranking_fulldata",
                          elpd_diff = "elpd.diff"),
           show_col_types = FALSE) %>%
  filter(col %in% cols) %>%
  mutate(elpd_diff = abs(elpd_diff) / sum(abs(elpd_diff)))
stopifnot(all(pull(df_elpd, col) == cols[-1]))

## Group MIBI-TOF granulomas with simulated GranSim granulomas by log-CFU.
##
## Trajectory plot of proximal GranSim simulatrions for each MIBI grauloma
## within bandwidth of MIBI log10-CFU chosen from Silverman's rule of thumb.
data(mibi_cells)
data(meta)
df_gs_traj_all <-
  read_csv(file_gs_traj,
           col_type = "cid",
           col_select = c(exp, time, TotMtbInGranScaledRing)) %>%
  mutate(exp = as.integer(exp),
         time = time / ticks_per_day) %>%
  right_join(select(df_gs, exp),
             by = "exp")
df_gs_traj <-
  df_gs_traj_all %>%
  filter(time == 11 * 7) %>%
  select(-time) %>%
  filter(TotMtbInGranScaledRing > 0)
df_mibi_traj <-
  mibi_cells %>%
  distinct(sample) %>%
  inner_join(meta, by = "sample") %>%
  select(sample, gran_cfu) %>%
  mutate(sample = fct_inorder(sample)) %>%
  filter(gran_cfu > 0)
df_mibi <-
  df_mibi %>%
  ## Remove the sterilizing granuloma.
  right_join(distinct(df_mibi_traj, sample),
             by = "sample")
bw <-
  df_mibi_traj %>%
  summarize(bw = stats::bw.nrd0(log10(gran_cfu))) %>%
  pull()
df_in_range <-
  mibi_cells %>%
  distinct(sample) %>%
  group_by(sample) %>%
  ## Duplicate all the GranSim experiments for each MIBI sample.
  reframe(exp = pull(df_gs_traj, exp),
          TotMtbInGranScaledRing = pull(df_gs_traj, TotMtbInGranScaledRing)) %>%
  ## Filter to +/- bandwidth.
  inner_join(df_mibi_traj %>%
               mutate(lb = 10 ^ (log10(gran_cfu) - bw),
                      ub = 10 ^ (log10(gran_cfu) + bw)),
             by = "sample") %>%
  group_by(sample) %>%
  filter(between(TotMtbInGranScaledRing, lb, ub)) %>%
  ungroup()
df_in_range %>%
  count(sample) %>%
  print(n = Inf)

## Pair granulomas before optimal transport only when they have no NAs.
stopifnot(df_in_range == drop_na(df_in_range))

n_replicates <- 10
df_pairs_reps <- vector("list", length = n_replicates)
df_in_range_reps <- df_in_range
for (i in seq_len(n_replicates)) {
  ## Prepare GranSim feature matrix with dimensions [F, L].
  mat_gs <-
    select(df_in_range_reps, exp) %>%
    left_join(df_gs,
              by = "exp") %>%
    select(-exp) %>%
    as.matrix() %>%
    `dimnames<-`(NULL) %>%
    t()
  ## Prepare MIBI-TOF feature matrix with dimensions [F, L].
  mat_mibi <-
    select(df_in_range_reps, sample) %>%
    left_join(df_mibi,
              by = "sample") %>%
    select(-sample) %>%
    as.matrix() %>%
    `dimnames<-`(NULL) %>%
    t()
  ## Calculate the cost vector.
  cost_vec <- c(pull(df_elpd, elpd_diff) %*% abs(mat_gs - mat_mibi))
  ## Run optimal transport.
  ##
  ## Methods equation 2 constraint of each MIBI-TOF granuloma having a single
  ## GranSim granuloma assigned.
  vec_gs_per_mibi <- count(df_in_range_reps, sample) %>% pull()
  stopifnot(sum(vec_gs_per_mibi) == nrow(df_in_range_reps))
  n_mibi <- length(vec_gs_per_mibi)
  n_combn <- sum(vec_gs_per_mibi)
  constraints_mibi <-
    spMatrix(nrow = n_mibi,
             ncol = n_combn,
             i = rep(seq_len(n_mibi),
                     times = vec_gs_per_mibi),
             j = seq_len(n_combn),
             x = rep(1, n_combn))
  ## Methods equation 3 constraint of each GranSim granuloma being assigned to
  ## a MIBI-TOF granuloma no more than once.
  df_gs_dup <-
    df_in_range_reps %>%
    mutate(j = row_number(),
           .before = 1) %>%
    left_join(count(df_in_range_reps, exp),
              by = "exp") %>%
    filter(n > 1) %>%
    group_by(exp) %>%
    mutate(i = cur_group_id())
  n_gs_dup <- max(pull(df_gs_dup, i))
  constraints_gs <-
    spMatrix(nrow = n_gs_dup,
             ncol = n_combn,
             i = pull(df_gs_dup, i),
             j = pull(df_gs_dup, j),
             x = rep(1, nrow(df_gs_dup)))
  ot <-
    Rglpk_solve_LP(obj = cost_vec,
                   mat = rbind(constraints_mibi,
                               constraints_gs),
                   dir = c(rep("==", nrow(constraints_mibi)),
                           rep("<=", nrow(constraints_gs))),
                   rhs = rep(1, nrow(constraints_mibi) + nrow(constraints_gs)),
                   types = "B",
                   max = FALSE)
  stopifnot(ot$status == 0L)
  ot_soln <- ot$solution
  stopifnot(sum(ot_soln) == n_mibi)
  df_pairs_reps[[i]] <-
    df_in_range_reps %>%
    filter(as.logical(ot_soln))
  ## All MIBI-TOF samples must be present.
  stopifnot(select(df_pairs_reps[[i]], sample) ==
              distinct(df_in_range_reps, sample))
  ## No GranSim exp should be repeated.
  stopifnot(pull(df_pairs_reps[[i]], exp) ==
              unique(pull(df_pairs_reps[[i]], exp)))
  ## Remove the matches from the simulations for the next OT iteration.
  exps_to_remove <- pull(df_pairs_reps[[i]], exp)
  n_exp <- function(df) nrow(distinct(df, exp))
  n_exp_before <- n_exp(df_in_range_reps)
  df_in_range_reps <-
    df_in_range_reps %>%
    filter(! exp %in% exps_to_remove)
  # Number of removed GranSim exp must be the same number as mibi.
  stopifnot(n_exp(df_in_range_reps) == n_exp_before - length(exps_to_remove))
  stopifnot(length(exps_to_remove) == n_mibi)
}

df_pairs <- bind_rows(df_pairs_reps, .id = "rep")
## All MIBI-TOF samples must be present.
stopifnot(distinct(df_pairs, sample) == distinct(df_in_range, sample))
## No GranSim exp should be repeated.
stopifnot(pull(distinct(df_pairs, exp)) == unique(pull(df_pairs, exp)))
## Each replicate should have a unique MIBI-TOF sample.
stopifnot(all(pull(count(df_pairs, sample, exp)) == 1L))

## Identify optimal granulomas among alternatives.
df_with_ot <-
  df_in_range %>%
  select(sample, exp) %>%
  left_join(df_pairs %>%
              select(rep, sample, exp),
            by = c("sample", "exp"))
```

```{r figure-06-ot-trajectories}
#| cache: true
#| dependson: -1
#| fig.width: 6.5
#| fig.height: 5.5
## Plot trajectories of the top-3 paired and all candidate granulomas.
relabel <- function(x) {
  str_c("DPs of ", str_replace(x, "sample", "Sample "))
}
n_colors <- 6L
colors <- palette.colors(n_colors + 1L, palette = "Okabe-Ito")[-1]
df_with_ot %>%
  inner_join(df_gs_traj_all,
             by = "exp",
             relationship = "many-to-many") %>%
  mutate(rep = ifelse(as.integer(rep) > 3L, NA, rep)) %>%
  rename(cfu = TotMtbInGranScaledRing) %>%
  mutate(log_cfu = log1p(cfu) / log1p(9),
         time = time / 7L,
         sample = fct_relabel(sample, relabel)) %>%
  group_by(sample, exp) %>%
  mutate(color_id =
           (cur_group_id() - 1L) %% n_colors %>%
           as.character() %>%
           fct(),
         .before = time) %>%
  ungroup() %>%
  ggplot(aes(time, cfu, group = exp, color = color_id)) +
  theme_bw() +
  theme(legend.direction = "horizontal",
        legend.position = "inside",
        legend.position.inside = c(1, 0.05),
        legend.justification = c("right", "center"),
        legend.title.position = "top",
        strip.text = element_text(size = 8)) +
  labs(x = "Week (post-infection)",
       y = "CFU counts",
       linetype = "Rank of digital partner") +
  guides(color = "none") +
  facet_wrap(~ sample) +
  geom_line() +
  geom_line(data =
              . %>%
              filter(! is.na(rep)),
            aes(linetype = rep,
                color = NULL)) +
  geom_point(data =
               df_mibi_traj %>%
               rename(cfu = gran_cfu) %>%
               mutate(sample = fct_relabel(sample, relabel),
                      time = 11),
             aes(group = NULL,
                 color = NULL)) +
  geom_text(data =
              df_in_range %>%
              count(sample) %>%
              mutate(sample = fct_relabel(sample, relabel),
                     label = sprintf("n = %s", format(n, big.mark = ","))),
            aes(group = NULL,
                color = NULL,
                label = label),
            x = -Inf,
            y = Inf,
            hjust = -0.05,
            vjust = 1.4,
            size = 3) +
  scale_color_manual(values = colors) +
  ## Remove padding around between data and axes.
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_log10(labels = scales::label_log())
```

```{r figure-07-ot-images}
#| cache: true
#| dependson: -2
## Plot images of experimental data and top-3 digitally paired granulomas.
##
## For each experimental sample, plot 3x3 partner images with the first row
## being experimental tissue (MIBI-TOF) and the each of the subsequent rows
## being timepoints from a paired granuloma simulation (weeks 4, 11, 22).

## Import TIFF images generated using the data-raw/merge.ijm ImageJ script.
dir_im_gransim <- file.path(dir_data, "2024-07-19-A-gr-50k", "img_stacks")
dir_im_mibi <- file.path(dir_data, "2024-07-19-A-gr-50k", "img_stacks_mibi")
## Add file paths to each dataset.
df_im <-
  df_with_ot %>%
  drop_na() %>%
  arrange(sample, rep)
df_im <-
  df_im %>%
  inner_join(distinct(df_im, sample) %>%
               group_by(sample) %>%
               mutate(path_mibi =
                        str_remove(sample, "sample") %>%
                        str_c("*/exp", ., "_*.tif") %>%
                        fs::dir_ls(dir_im_mibi, glob = .)),
             by = "sample") %>%
  inner_join(distinct(df_im, exp) %>%
               group_by(exp) %>%
               reframe(path_gransim =
                         str_c("*/exp", exp, "_*.tif") %>%
                         fs::dir_ls(dir_im_gransim, glob = .) %>%
                         gtools::mixedsort()),
             by = "exp")
## The uncompressed TIFF images are relatively large, so only read and save one
## image set at a time.
##
#' Create composite color image from the 3 channels.
im_color <- function(path,
                     colors = c("#0000ff", "#00ff00", "#ffffff"),
                     pixels_to_crop = 100,
                     add_scalebar = FALSE,
                     scalebar_microns = 500) {
  im <- suppressWarnings(readImage(path))
  ## Create composite image.
  im_c <- imageData(im)
  for (ch in seq_len(numberOfFrames(im))) {
    im_ch <- im_c[, , ch]
    mask <- im_ch > 0
    im_ch[mask] <- colors[ch]
    im_ch[! mask] <- "#000000"
    im_c[, , ch] <- im_ch
  }
  ## Creates additional color channel dimension [x, y, ch, color].
  im_c <- Image(im_c)
  ## Merge channels.
  im_max <- rowSums(imageData(im_c), dims = 3)
  ## Crop image
  by <- pixels_to_crop
  idx <- (1 + by):(dim(im_max)[1] - by)
  im_max <- im_max[idx, idx, ]
  if (add_scalebar) {
    ## Add 500 micron scale bar to bottom right.
    width <- scalebar_microns / 20
    height <- 2
    offset <- 5
    x <- dim(im_max)[1]
    ## Sometimes the scale bar overlaps with white pixels, so invert
    ## the color instead of setting to white.
    im_max[x - (offset + 1:width), x - (offset + 1:height), ] <-
      1 - im_max[x - (offset + 1:width), x - (offset + 1:height), ]
  }
  ## Return final color image.
  Image(im_max, colormode = "Color")
}
dir_fig_suppl <- knitr::opts_chunk$get("fig.path")
n_groups <-
  df_im %>%
  group_by(sample) %>%
  n_groups()
for (i in seq_len(n_groups)) {
  df <-
    df_im %>%
    group_by(sample) %>%
    nest() %>%
    ungroup() %>%
    slice(i) %>%
    unnest(data)
  path_mibi <-
    df %>%
    distinct(path_mibi) %>%
    pull()
  ## Read in the MIBI image.
  im_mibi <- im_color(path_mibi, add_scalebar = TRUE)
  ## Pad the MIBI image on the left and right sides.
  im_blank <- im_mibi
  imageData(im_blank) <- array(1L,
                               dim = dim(im_blank),
                               dimnames = dimnames(im_blank))
  ims_mibi <- combine(im_blank, im_mibi, im_blank)
  ## Read in the GranSim images.
  path_gransim <-
    df %>%
    distinct(path_gransim) %>%
    pull()
  ims_gransim <- combine(lapply(path_gransim, im_color))
  ims <- combine(ims_mibi, ims_gransim)
  display(ims,
          method = "raster",
          all = TRUE,
          nx = 3L)
  ## Add subfigure labels.
  subfig <- c(NA, LETTERS[1], NA, LETTERS[2:10])
  ncol <- 3
  for (i in seq_along(subfig)) {
    if (! is.na(subfig[i])) {
      row <- (i - 1) %/% ncol + 1
      col <- (i - 1) %% ncol + 1
      offset <- 10
      mul <- dim(ims)[1]
      x <- mul * (row - 1) + offset
      y <- mul * (col - 1) + offset
      text(subfig[i],
           x = y,
           y = x,
           col = "white")
    }
  }
  path_fig_suppl <-
    fs::path(dir_fig_suppl,
             sprintf("%s.png",
                     df %>%
                       distinct(sample) %>%
                       pull()))
  dev.print(png,
            path_fig_suppl,
            width = 300,
            height = 400,
            units = "px",
            res = 100)
  message("Saved ", path_fig_suppl)
}
```

```{r session-info}
#| cache: false
sessionInfo()
```
